---
title: "Understanding the central limit theorem"
output: 
  flexdashboard::flex_dashboard:
    orientation: columns
    vertical_layout: fill
    theme: simplex
runtime: shiny
---

```{r setup, include=FALSE}

# load relevant libraries
library(flexdashboard)
library(shiny)
library(dplyr)
library(tidyr)
library(ggplot2)

#' @title: sample_clt()
#' @description: function to sample from a distribution and estimate the 
#' sample mean, standard deviation, n and standard error.
#' @param pop_data - vector of data-points from the population
#' @param n - size of the sample to be taken
#' @param n_samples - number of indepedent samples of size n to take

sample_clt <- function(pop_data, n, n_samples) {
  
  # test input data
  assertthat::assert_that(assertthat::is.count(n), 
                          msg = "n must be an integer")
  
  assertthat::assert_that(assertthat::is.count(n_samples), 
                         msg = "n_samples must be an integer")
  
  assertthat::assert_that(n < length(pop_data),
                          msg = "size of the sample cannot be greater than
                          the number of samples in the population distribution")
  
  dist_clt <- vector("list", length = length(1:n_samples))
  for (i in seq_along(1:n_samples)) {
    
    samp_dat <- sample(pop_data, size = n, replace = FALSE, prob = NULL)
    m_dist <- mean(samp_dat)
    sd_dist <- sd(samp_dat)
    n_dist <- length(samp_dat)
    se_dist <- sd(samp_dat)/sqrt(length(samp_dat))
    dist_clt[[i]] <- dplyr::bind_cols(sample_mean = m_dist,
                                      sample_sd = sd_dist,
                                      sample_n = n_dist,
                                      sample_se = se_dist)
  }
  
  # combine the independent samples
  dist_clt <- dplyr::bind_rows(dist_clt)
  
  # combine the different samples
  dplyr::bind_cols( dplyr::tibble(sample_size = as.character(rep(n, times = n_samples))),
                    dplyr::select(dist_clt, sample_mean, sample_sd, sample_n, sample_se))
  
}

#' @title process_hist()
#' @description function to process the histogram plotting that this dashboard
#' repeats several times
#' @param vec - vector of numeric data
#' @param min_x - minimum limit on the x-axis
#' @param min_y - minimum limit on the y-axis
#' @param mean_y - label for the mean
#' @param sd_y - label for the standard deviation
#' @param add_labels - whether to add the mean and sd labels
process_hist <- function(vec, min_x, max_x, mean_lab, sd_lab, add_labels,
                         sca_yaxis = 0.065,
                         sca_mean = 0.07,
                         sca_seg = 0.16,
                         sca_sd = 0.22) {
  
  # add the distribution to a table
  x <- dplyr::tibble(y = vec)
  
  # get the length of the vector
  n_vec <- length(vec)
  
  # calculating the Sturges bins
  breaks <- pretty(range(x$y), n = nclass.Sturges(x$y), min.n = 1)
  
  # classify each observation into a bin
  group_j <- vector(length = nrow(x))
  for(j in 1:nrow(x)) {
    group_i <- vector(length = length(breaks)-1)
    for(i in 1:length(breaks-1)) {
      group_i[i] <- (x$y[j] > breaks[i]) & (x$y[j] < breaks[i+1])
      }
    group_j[j] <- which(group_i == TRUE) 
    }
  
  # add the groups to the table
  x$z <- group_j
  
  # calculate the mean and standard deviation
  mean_x <- mean(x$y)
  sd_x <- sd(x$y)
  
  # count per break
  x_sum <- 
    x |>
    dplyr::group_by(z) |>
    dplyr::summarise(midpoint = mean(y),
                     count = n())
  
  # get the maximum group count
  ymax <- max(x_sum$count)
  
  # plot this as a histogram
  p1 <- 
    ggplot() +
    geom_histogram(data = x, mapping = aes(x = y), 
                   breaks = breaks, colour = "grey") +
    geom_segment(mapping = aes(x = mean_x, xend = mean_x, 
                               y = 0, yend = ymax),
                 colour = "red", linewidth = 1) +
    geom_rect(mapping = aes(xmin = mean_x-sd_x, xmax = mean_x+sd_x, ymin = 0, ymax = ymax),
              alpha = 0.1, fill = "red") +
    scale_y_continuous(expand = c(0, 0), limits = c(0, ymax + (sca_yaxis*n_vec) )) +
    scale_x_continuous(limits = c(min_x, max_x)) +
    xlab(NULL) +
    ylab(NULL) +
    theme_meta() +
    theme(axis.text.y = element_blank(),
          axis.ticks.y = element_blank())
  
  if(add_labels) {
    p1 <- 
      p1 +
      annotate(geom = "text", x = mean_x, y =  (ymax + (ymax*sca_mean)), label = mean_lab,
             size = 6, colour = "red") +
      geom_segment(mapping = aes(x = mean_x, xend = mean_x + sd_x, 
                               y =  (ymax + (ymax*sca_seg)), yend =  (ymax + (ymax*sca_seg))),
                   colour = "red", linewidth = 0.5) +
      annotate(geom = "text", x = ((mean_x) + (sd_x)/2), y = (ymax + (ymax*sca_sd)), label = sd_lab,
               size = 6, colour = "red")
  }
  
  p1
  
}

#' @title theme_meta()
#' @description customised plotting theme to equalise the formatting of
#'  all plots plotted with ggplot2
#' @author James G. Hagan (james_hagan(at)outlook.com)
theme_meta <- function(base_size = 12, base_family = "") {
  theme(
    panel.background = element_rect(fill = "white"),
    panel.border = element_blank(),
    axis.line.x = element_line(color = "black", linewidth = 0.2),
    axis.line.y = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.ticks.length = unit(-0.16, "cm"),
    axis.title.x = element_text(
      colour = "black",
      size = 12,
      face = "plain",
      margin = margin(5, 0, 0, 0, "pt")
    ),
    axis.title.y = element_text(
      colour = "black",
      size = 12,
      face = "plain",
      margin = margin(0, 5, 0, 0, "pt")
    ),
    axis.text.x = element_text(
      colour = "black",
      size = 11,
      face = "plain",
      margin = margin(10, 0, 0, 0, "pt")
    ),
    axis.text.y = element_text(
      colour = "black",
      size = 11,
      face = "plain",
      margin = margin(0, 10, 0, 0, "pt")
    ),
    axis.ticks.x = element_line(colour = "black", size = 0.4),
    axis.ticks.y = element_line(colour = "black", size = 0.4),
    legend.text = element_text(colour = "black", size = 10, face = "plain"),
    legend.title = element_text(colour = "black", size = 10, face = "plain")
  )
}

```

Sidebar {.sidebar data-width=375}
======================================================================

```{r}
# population distribution parameters
numericInput("pop_size", "Population size (N)", value = 10000, min = 1000, max = 100000)
numericInput("shape", "Shape parameter", value = 20, min = 0, max = 15, step = 0.5)
numericInput("rate", "Rate parameter", value = 1, min = 1, max = 20, step = 0.5)

# size of the sample to take
sliderInput("n", "Sample size (n)", value = 30, min = 1, max = 200, step = 1)

# simulate the population distribution
popx <- 
  reactive(rgamma(n = input$pop_size, shape = input$shape, rate = input$rate))

min_popx <- reactive(min(popx()))
max_popx <- reactive(max(popx()))
n_popx <- reactive(length(popx()))

# take independent samples from the population distribution
samplex <- reactive(sample_clt(pop_data = popx(), n = input$n, n_samples = 1000))
           
```

The Central Limit Theorem states that the distribution of sample means approximates a normal distribution with a mean equal to the true population mean $(\mu)$ and a variance equal to the true population variance $(\sigma^2)$ divided by the sample size (n): $(\frac{\sigma^2}{n})$. The CLT is one of the most fundamental theorems in statistics and it allows us to do many magical things like calculate confidence intervals which is demonstrated in this application.

Using this application, we can simulate a population from a Gamma distribution (**a**). This population has some true mean $(\mu)$ and some standard deviation $(\sigma)$. Then, we can draw a random sample from this population distribution of size n which gives us the sample mean $(\bar{x})$ and some sample standard deviation (s), (**b**). If we continue to take random samples of size and calculate the sample mean $(\bar{x})$ for each random sample, eventually we can construct a distribution of sample means (**c**). This distribution of sample means (i.e. $\bar{x}$ values) is the *sampling distribution of the mean*. The amazing thing is that, irrespective of the population distribution, the sampling distribution tends to approximate a normal distribution with a mean equal to the population mean $(\mu)$. Try varying the shape and rate parameters and see how stable the shape of sampling distribution is.

We can then exploit this property to calculate confidence intervals. In **d**, the 95% confidence interval is calculated for each random sample. Red intervals contain the true population mean $(\mu)$. Black intervals do not. As you can see, in most cases, the 95% confidence interval, which is calculated using the sample mean $(\bar{x})$ and the sample standard deviation (s) of each sample, contain the true population mean. In fact, the actual number is almost always very close to 95% which is where the interpretation of the 95% confidence interval comes from. Specifically, the interpretation of the 95% confidence interval is that, if we were to conduct the same experiment (in this case take random samples from the population) say 1000 times, in 95% of those experiments, the calculated 95% confidence would contain the true population mean. Note that, even with very skewed population distributions, this tends to remain true. This is the power of the CLT.

Application author: **James Hagan**

Body {.hidden}
======================================================================

Column {data-width=100}
-------------------------------------------------------------------------------

### a. Population distribution (N)

```{r}

renderPlot({
  
  # get the population-level histogram
  process_hist(vec = popx(), min_x = min_popx(), max_x = max_popx(),
               mean_lab = expression(mu), sd_lab = expression(sigma),
               add_labels = TRUE,
               sca_yaxis = 0.06,
               sca_mean = 0.07,
               sca_seg = 0.16,
               sca_sd = 0.22)
  
}

)

```

### b. One random sample from the population (i.e. size = n)

```{r}

renderPlot({
  
  # get the population-level histogram
  process_hist(vec = sample(popx(), size = input$n, replace = FALSE), 
               min_x = min_popx(), max_x = max_popx(),
               mean_lab = expression(bar("x")), sd_lab = "s",
               add_labels = TRUE,
               sca_yaxis = 0.15,
               sca_mean = 0.08,
               sca_seg = 0.17,
               sca_sd = 0.26)
  
}

)

```


### c. Sampling distribution (i.e. distribution of x-bar values from 1000 seperate sample of size = n)

```{r}
renderPlot({
  
  # get the population-level histogram
  process_hist(vec = samplex()$sample_mean, 
               min_x = min_popx(), max_x = max_popx(),
               add_labels = FALSE,
               sca_yaxis = 0.01)
  
}

)
```


Column {data-width=90}
-------------------------------------------------------------------------------

### d. Confidence intervals (95%) for each random sample

```{r}

renderPlot({
  
  # calculate confidence intervals
  conf_df <- dplyr::bind_cols(tibble(id = 1:nrow(samplex()), samplex()))
  
  # calculate the 95% confidence interval for each random sample
  
  # lower interval
  conf_df$ci_low <- 
    with(conf_df, 
         sample_mean + (qt(p = c(0.025), df = conf_df$sample_n)*sample_se))
  
  # upper interval
  conf_df$ci_upp <- 
    with(conf_df, 
         sample_mean + (qt(p = c(0.975), df = conf_df$sample_n)*sample_se))
  
  # get the population mean
  mean_x <- mean(popx())
  
   # check which confidence intervals contain the true mean
  conf_df$mean_within <- 
    with(conf_df,
         (mean_x > ci_low) & (mean_x < ci_upp) )
  
  # take N random samples
  N <- 300
  conf_df_sub <- conf_df[sample(x = 1:nrow(conf_df), size = N, replace = FALSE),]
  conf_df_sub$id <- 1:nrow(conf_df_sub)
  
  # calculate percentage within
  perc_within <- (sum(conf_df$mean_within)/nrow(conf_df))*100
  
  ggplot(data = conf_df_sub) +
  geom_segment(mapping = aes(x = ci_low, xend = ci_upp, y = id, yend = id,
                             colour = mean_within), linewidth = 0.35, alpha = 0.8) +
  scale_colour_manual(values = c("black", "red")) +
  geom_segment(mapping = aes(x = mean_x, xend = mean_x, 
                             y = 0, yend = max(id)+5), colour = "red", linewidth = 0.5) +
  annotate(geom = "text", x = mean_x, y =  max(conf_df_sub$id)+15, label = expression(mu),
           size = 6, colour = "red") +
  annotate(geom = "text", x = mean_x, y =  max(conf_df_sub$id)+15, hjust = -0.5,
           label = bquote(.(perc_within)*"% contain" ~ mu),
           size = 5, colour = "red") +
  scale_y_continuous(expand = expansion(mult = c(0.01, 0)), limits = c(0, N+20)) +
  theme_meta() +
  xlab(NULL) +
  ylab(NULL) +
  theme_meta() +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        legend.position = "none")
  
}) 

```







